Results of two multi-chord stellar occultations by dwarf planet (1) Ceres

### 2017-08-16 Ceres/paper-Ceres-revised/paper-Ceres.tex
mod 2017 ago16

Ceres is the sole example of a dwarf planet in the inner Solar System. Far from being mere taxonomic information, this suggests the great impact its study can have on the understanding of planetary formation and evolution of the Solar System. Indeed, it was proposed that Ceres' origin could have been as a transneptunian object \citep{McKinnon12}, later scattered to the Main Belt due to the giant planets' migration predicted by the `Nice Model' \citep{Gomes05}. Even if it was formed close to its current location, the dynamical history of the Solar System must have let its signatures on Ceres. These could include not only the late heavy bombardment features that might exist on its surface, but also the makeup of its volatiles, which could have been transported from the outer region.

Since the 1970's it has been speculated that Ceres could contain water ice, which was recently verified \citep{Kuppers2014}. Although the water regime on this object is still unknown, some internal structure models suggest the existence of a water ice -- or even a liquid water -- layer \citep{CastilloR2011}. Yet the very question of whether Ceres underwent differentiation is open and, on the assumption of an affirmative answer, it is natural to ask if it ever had tectonics, what its geological evolution was, and if it is still active. Inarguably NASA's \textit{Dawn} mission \citep{Russell2004}, which is currently orbiting the dwarf planet, will shed light on several open issues concerning Ceres.

@@

From observations with adaptive optics spanning a 10-year period, \cite{Drummond2014} determined the position of Ceres' polar axis within an error of only 3$\degr$:
\begin{equation}
%\left\{ 
%\begin{array}{l}
\alpha = (287 \pm 3) \degr, \quad %\\
\delta = (+64 \pm 3) \degr,
%\end{array} \right .
\label{Eq: Pole}
\end{equation}
%
in equatorial J2000 coordinates. Together with Ceres' ephemeris at the moment of the occultation, this corresponds to the polar aspect angle $\zeta = 86.1\degr$, which is very close to an equator-on geometry. Hence, we expect true figures to be similar to apparent ones.

The knowledge of Ceres' pole not only allows the determination of its polar aspect, it suffices to set its position angle. Therefore Eq.~\ref{Eq: Pole} may act as a constraint for $P$, and a second solution can be obtained by probing the parameter space with the restriction that the position angle is confined to the range that follows from Eq.~\ref{Eq: Pole}. We call this the `pole-constrained solution'.

@@

The 2010 event, for example, had only seven contacts; none the less, they were well distributed over Ceres' disc (see Fig. \ref{Fig:Ceres-2010-body}) acting as a constraint to its shape. On the other hand, the 2013 event had five more exploitable contacts, but they were concentrated in certain regions of the body. In particular, the absence of chords close to Ceres' south pole made its oblateness worse determined here than in the 2010 event.

@@

The few chords of the 2010 occultation could themselves only constraint the position angle of the object to a uncertainty of $10\degr$. This uncertainty was reduced by a factor of two in the 2013 event, approaching --and verifying-- the result predicted by the work of \cite{Drummond2014}. As was shown, using the coordinates of Ceres' polar axis to limit the position angle was not an efficient procedure in the 2013 event, in the sense that it did not result in significant changes in the parameters obtained in the nominal solution (see Table \ref{Tab: results}).

On the other hand, constrain the position angle on the 2010 occultation was proved to reduce the error bars of the other parameters (disregarding oblateness). Moreover, this procedure resulted in an excellent agreement between the equatorial radius figures of both events.

@@

As mentioned in the Introduction, the 1984 event \citep{Millis1987} is the only occultation data to which we can compare our result. The measured diameters do not agree, ours being larger by 2-sigma. It is difficult to state for sure the reasons for this divergence since we had no access to the original 1984 data or reduction methodology. We note, however, that in this occultation a variation of 0.5 s in the contact times correspond to almost 7 km on Ceres' limb, which is on the same order of the residuals of their elliptical fit. Therefore, in order to properly compare their result to ours it would be necessary to reduce their light curves with the same methodology used in the present work.

@@

\begin{table}
  \caption{Ceres' equatorial diameter and oblateness \label{Tab: Ceres-final}}
  \begin{centering}
  \begin{tabular}{@{}lccc}
  \hline
     Eq. diameter (km) & Oblateness & Method & Ref. \\
\hline
$972 \pm 6$  & $0.08  \pm 0.03$  & Occultation & 1\\
$967 \pm 10$ & $0.078 \pm 0.015$ & Keck+VTL    & 2 \\
$959 \pm 5$  & $0.074 \pm 0.007$ & Keck        & 3\\
$975 \pm 4$  & $0.067 \pm 0.005$ & HST         & 4\\
$959 \pm 5$  & $0.05  \pm 0.01$  & Occultation & 5\\
\hline
\end{tabular}
\par\end{centering}
\textbf{References.} 1: Present work. 2: \cite{Drummond2014}. 3: \cite{Carry2008}. 4: \cite{Thomas2005}. 5: \cite{Millis1987}.
\end{table}

@@

ARGJ thanks the financial support of CAPES. BLG thanks CNPq. F.B.R. acknowledges PAPDRJ-FAPERJ/CAPES E-43/2013 number 144997, E-26/101.375/2014 and CDFB-CAPES/Brazil. M.A. acknowledges CNPq grants 473002/2013-2, 482080/2009-4 and 312394/2014-4, and FAPERJ grant 111.488/2013. RVM thanks grants: CNPq-306885/2013, Capes/Cofecub-2506/2015, Faperj/PAPDRJ-45/2013. J.I.B. Camargo acknowledges CNPq for a PQ2 fellowship (process number 308489/2013-6). We also acknowledge Steve Preston for the predictions of the occultations.

>>>>

Ceres is the sole example of a dwarf planet in the inner Solar System. Far from being mere taxonomic information, this suggests the great impact its study can have on the understanding of planetary formation and evolution of the Solar System. Indeed, it was proposed that Ceres' origin could have been as a transneptunian object \citep{McKinnon12}, later scattered to the Main Belt due to the giant planets' migration predicted by the `Nice Model' \citep{Gomes05}. Even if it was formed close to its current location, the dynamical history of the Solar System must have left its signatures on Ceres. These could include not only the late heavy bombardment features that might exist on its surface, but also the makeup of its volatiles, which could have been transported from the outer region.

Since the 1970's it has been speculated that Ceres could contain water ice, which was recently verified \citep{Kuppers2014}. Although the water regime on this object is still unknown, some internal structure models suggest the existence of a water vapour -- or even a liquid water -- layer \citep{CastilloR2011}. Yet the very question of whether Ceres underwent differentiation is open and, on the assumption of an affirmative answer, it is natural to ask if it ever had tectonics, what its geological evolution was, and if it is still active. Inarguably NASA's \textit{Dawn} mission \citep{Russell2004}, which is currently orbiting the dwarf planet, will shed light on several open issues concerning Ceres.

@@

\textbf{It is possible to derive the pole position $(\zeta, P)$ from the coordinates of Ceres' pole $(\alpha_{p}, \delta_{p})$ and its ephemeris $(\alpha, \delta)$ via}
%
\begin{equation}
\left\{
\begin{array}{l l}
\cos(\zeta)= -\sin(\delta_{p})\sin(\delta) - \cos(\delta_{p})\cos(\delta)\cos(\alpha_{p}-\alpha)\\
\tan(P) = \dfrac{\sin(\alpha_{p}-\alpha)} {\tan(\delta_{p})\cos(\delta) - \sin(\delta)\cos(\alpha_{p}-\alpha)},\\
%\cos(B).\cos(P)= \sin(\delta_{p}).\cos(\delta) - \cos(\delta_{p}).\sin(\delta).\cos(\alpha_{p}-\alpha)\\
%\cos(B).\sin(P)=\cos(\delta_{p}).\sin(\alpha_{p}-\alpha)\\
%\zeta = 90\degr - B
\end{array}
\right.
\label{Eq: pole-position}
\end{equation}
%
\textbf{where $\zeta = 0\degr$ and $\zeta=90\degr$ respectively correspond to a pole-on and an equator-on geometry.}

From observations with adaptive optics spanning a 10-year period, \cite{Drummond2014} determined the position of Ceres' polar axis within an error of only 3$\degr$:
\begin{equation}
%\left\{ 
%\begin{array}{l}
\alpha_{p} = (287 \pm 3) \degr, \quad %\\
\delta_{p} = (+64 \pm 3) \degr,
%\end{array} \right .
\label{Eq: Pole}
\end{equation}
%
in equatorial J2000 coordinates. Together with Ceres' ephemeris at the moment of the occultation, this corresponds to the polar aspect angle $\zeta = 86.1\degr$, which is very close to an equator-on geometry. Hence, we expect true figures to be similar to apparent ones.

The knowledge of Ceres' pole not only allows the determination of its polar aspect, it suffices to set its position angle. Therefore \textbf{Eqs. \ref{Eq: pole-position} and \ref{Eq: Pole}} may act as a constraint for $P$, and a second solution can be obtained by probing the parameter space with the restriction that the position angle is confined to the range that follows from Eq. \ref{Eq: Pole}. We call this the `pole-constrained solution'.

@@

The 2010 event, for example, had only seven contacts; none the less, they were well distributed over Ceres' disc (see Fig. \ref{Fig:Ceres-2010-body}) acting as a constraint to its shape. On the other hand, the 2013 event had five more exploitable contacts, but they were concentrated in certain regions of the body. In particular, the absence of chords close to Ceres' south pole made its oblateness less well determined here than in the 2010 event.

@@

The few chords of the 2010 occultation could themselves only \textbf{constrain} the position angle of the object to a uncertainty of $10\degr$. This uncertainty was reduced by a factor of two in the 2013 event, approaching --and verifying-- the result predicted by the work of \cite{Drummond2014}. As was shown, using the coordinates of Ceres' polar axis to limit the position angle was not an efficient procedure in the 2013 event, in the sense that it did not result in significant changes in the parameters obtained in the nominal solution (see Table \ref{Tab: results}).

On the other hand, \textbf{constraining} the position angle on the 2010 occultation was proved to reduce the error bars of the other parameters (disregarding oblateness). Moreover, this procedure resulted in an excellent agreement between the equatorial radius figures of both events.

@@

As mentioned in the Introduction, the 1984 event \citep{Millis1987} is the only occultation data to which we can compare our result. The measured diameters do not agree, ours being larger by 2-sigma. \textbf{It is difficult to state for sure the reasons for this divergence. One way to clarify the issue is to redetermine the immersion and emersion instants from the original light curves using the same methodology presented in our work. Note that in this occultation a variation of 0.5 s in the contact times correspond to almost 7 km on Ceres' limb, which is on the same order of the residuals of their elliptical fit. However, we could not do that, as we had no access to the original 1984 light curve data. Moreover, we have no description of how the immersion and emersion instants of the chords were actually determined from these light curves in \cite{Millis1987}.}

@@

\begin{table}
  \caption{Ceres' equatorial diameter and oblateness \label{Tab: Ceres-final}}
  \begin{centering}
  \begin{tabular}{@{}lccc}
  \hline
     Eq. diameter (km) & Oblateness & Method & Ref. \\
\hline
$972 \pm 6$  & $0.08  \pm 0.03$  & Occultation & 1\\
$967 \pm 10$ & $0.078 \pm 0.015$ & Keck+VTL    & 2\\
$959 \pm 5$  & $0.074 \pm 0.007$ & Keck        & 3\\
$975 \pm 4$  & $0.067 \pm 0.005$ & HST         & 4\\
$959 \pm 5$  & $0.05  \pm 0.01$  & Occultation & 5\\
\hline
\end{tabular}
\par\end{centering}
\textbf{References.} 1: Present work. 2: \cite{Drummond2014}. 3: \cite{Carry2008}. 4: \cite{Thomas2005}. 5: \cite{Millis1987}.
\end{table}

@@

ARGJ thanks the financial support of CAPES. BLG thanks CNPq. F.B.R. acknowledges PAPDRJ-FAPERJ/CAPES E-43/2013 number 144997, E-26/101.375/2014 and CDFB-CAPES/Brazil. M.A. acknowledges CNPq grants 473002/2013-2, 482080/2009-4 and 312394/2014-4, and FAPERJ grant 111.488/2013. RVM thanks grants: CNPq-306885/2013, Capes/Cofecub-2506/2015, Faperj/PAPDRJ-45/2013. J.I.B. Camargo acknowledges CNPq for a PQ2 fellowship (process number 308489/2013-6). We also acknowledge Steve Preston for the predictions of the occultations \textbf{and the referee Lawrence H. Wasserman (Lowell Observatory) for his contributions to improve the text.}
